---
title: "Measurement Error"
author: "Gabriel Mesevage"
format:
  revealjs:
    slide-number: c/t
    chalkboard:
      buttons: true
editor:
  markdown:
    wrap: sentence
bibliography: references.bib
---

## Today's plan {.smaller}

1.  Classical measurement error and why averaging works
2.  Systematic but uncorrelated bias: we can still recover $\beta$
3.  Non-classical (correlated) bias: destroys inference
4.  Measurement error in the explanatory variable: attenuation bias
5.  Application: population data, Nunn & Qian, and Guinnane's critique

## What is classical measurement error? {.smaller}

-   Any measurement of a true quantity $\tau^*$ will contain some error

-   We write the measured value as $\tau_i = \tau^* + \epsilon_i$

-   **Classical** measurement error has three properties:

    1.  **Mean-zero**: errors are symmetric — as often too high as too low, $E[\epsilon_i] = 0$
    2.  **Uncorrelated with the true value**: big things are not measured with bigger errors
    3.  **Uncorrelated with other errors**: e.g. in a regression

-   Under these conditions, averaging works:

$$\text{Avg}(\tau_i) = \frac{1}{N}\sum_i \tau_i = \tau^* + \underbrace{\text{Avg}(\epsilon_i)}_{\to\, 0} \xrightarrow{N \to \infty} \tau^*$$

-   **Key result**: with enough measurements, the errors cancel and we recover the truth

## The experiment — round 1 {.smaller}

I will toss a book in the air.
Time how long it stays in the air (seconds, to 2 decimal places) and enter your measurement in the spreadsheet.

::: {style="text-align: center; margin-top: 0.5em;"}
```{r}
#| echo: false
#| fig-height: 3.8
#| fig-width: 3.8
#| message: false
#| warning: false
library(qrcode)
qr <- qr_code("https://tinyurl.com/253ndpxe")
plot(qr)
```

<https://tinyurl.com/253ndpxe>
:::

## What is systematic bias? {.smaller}

-   Classical measurement error assumes errors are mean-zero — but what if they are not?

-   **Systematic bias**: the average error is not zero, $E[\epsilon_i] = b \neq 0$

-   Example: when timing the book-toss, students may be systematically slow to hit stop — over-estimating time in the air on average

-   This is a **constant bias** — it shifts every measurement by the same amount

-   The good news: if the bias does not depend on what you are measuring, it can still be harmless for many purposes

-   Specifically, if the bias is the **same across all groups** being compared, it cancels in the comparison — as we will now show

## Systematic bias in the experiment {.smaller}

-   Let $D_i = 1$ if the toss is right-handed, $D_i = 0$ if left-handed
-   Suppose the bias $b$ is the same whichever hand I use:

$$\tau_i = D_i \tau^R + (1-D_i)\tau^L + b + \epsilon_i$$

Rearranging step by step:

$$\tau_i = D_i \tau^R + \tau^L - D_i \tau^L + b + \epsilon_i$$

$$\tau_i = \underbrace{\tau^L + b}_{\alpha} + \underbrace{(\tau^R - \tau^L)}_{\beta}\, D_i + \epsilon_i$$

-   The regression coefficient $\beta$ still estimates the **true difference** between right and left-handed tosses

-   The bias is absorbed entirely into the intercept $\alpha$

-   **Intuition**: if you make the same mistake measuring both groups, the mistake cancels when you take the difference

## The experiment — round 2 {.smaller}

The data from round 1 (right-hand toss) is still in the spreadsheet.
Now open the sheet and **add a left-hand toss only** — record the time and mark which hand.

::: {style="text-align: center; margin-top: 0.5em;"}
```{r}
#| echo: false
#| fig-height: 3.8
#| fig-width: 3.8
#| message: false
#| warning: false
library(qrcode)
qr <- qr_code("https://tinyurl.com/253ndpxe")
plot(qr)
```

<https://tinyurl.com/253ndpxe>
:::

## Non-classical bias: the algebra {.smaller}

-   Now suppose the bias is **different** for the two groups: $b^R$ for right-hand, $b^L$ for left-hand
-   The bias is correlated with the treatment $D_i$

Starting from:

$$\tau_i = D_i(\tau^R + b^R) + (1-D_i)(\tau^L + b^L) + \epsilon_i$$

Expanding:

$$\tau_i = D_i(\tau^R + b^R) + (\tau^L + b^L) - D_i(\tau^L + b^L) + \epsilon_i$$

$$\tau_i = \underbrace{(\tau^L + b^L)}_{\alpha} + \underbrace{[(\tau^R - \tau^L) + (b^R - b^L)]}_{\beta}\, D_i + \epsilon_i$$

-   The slope $\beta$ now **mixes together** the true difference $(\tau^R - \tau^L)$ and the bias difference $(b^R - b^L)$

## Non-classical bias: what it means {.smaller}

-   We cannot separate the true difference from the bias difference — they are bundled into a single number $\hat\beta$

-   A finding of $\hat\beta = 1$ second is consistent with **any** of:

    -   A real 1-second gap, zero bias difference
    -   No real gap, a 1-second bias difference
    -   A 2-second real gap offset by a −1-second bias difference
    -   … and infinitely many other combinations

-   Sometimes context or logic lets us put reasonable bounds on the biases — but not in general

-   **Lesson**: measurement error that is correlated with your explanatory variable destroys inference — you cannot recover the true relationship

## Measurement error in the explanatory variable {.smaller}

-   So far the error has been in the **outcome** $\tau_i$.
    What if the error is in the **explanatory variable**?

-   True model: $y_i = \alpha + \beta x^*_i + \epsilon_i$, but we observe $x_i = x^*_i + u_i$

-   Assume **classical** $u_i$: mean-zero, uncorrelated with $x^*_i$ and $\epsilon_i$

-   Recall $\hat\beta = \text{Cov}(x_i, y_i) / \text{Var}(x_i)$.
    Substituting:

$$\text{Cov}(x_i, y_i) = \text{Cov}(x^*_i + u_i,\; \beta x^*_i + \epsilon_i) = \beta\,\text{Var}(x^*_i)$$

$$\text{Var}(x_i) = \text{Var}(x^*_i) + \text{Var}(u_i)$$

Therefore:

$$\hat\beta = \beta \cdot \frac{\text{Var}(x^*_i)}{\text{Var}(x^*_i) + \text{Var}(u_i)}$$

## Attenuation bias — implications {.smaller}

$$\hat\beta = \beta \cdot \underbrace{\frac{\text{Var}(x^*_i)}{\text{Var}(x^*_i) + \text{Var}(u_i)}}_{\text{between 0 and 1}}$$

-   The fraction is always between 0 and 1: **attenuation bias** — $\hat\beta$ is pushed toward zero

-   The larger the noise variance $\text{Var}(u_i)$ relative to the signal, the closer $\hat\beta$ gets to zero

-   A **constant** error (zero variance) causes no bias — it is only *variation* in the error that attenuates

-   You will often see scholars argue: "even if there is measurement error, it just makes us under-estimate the relationship" — this is true only for **classical** error

-   If error in $x$ is non-classical, the bias is unknown in sign and magnitude

-   Moreover: ME in $x$ corrupts estimates of all other coefficients in a multiple regression, even if their own regressors are perfectly measured

## Attenuation bias — visualised {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| message: false
#| warning: false
library(ggplot2)
library(patchwork)

set.seed(42)
n <- 200
x_true <- rnorm(n, 0, 1)
y <- 1.5 * x_true + rnorm(n, 0, 0.5)
x_lim <- c(-13, 13)

make_plot <- function(noise_sd, label) {
  x_obs <- x_true + rnorm(n, 0, noise_sd)
  df <- data.frame(x = x_obs, y = y)
  b <- round(coef(lm(y ~ x, data = df))[2], 2)
  ggplot(df, aes(x, y)) +
    geom_point(alpha = 0.4, size = 0.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
    coord_cartesian(xlim = x_lim) +
    labs(title = label, subtitle = paste0("\u03b2\u0302 = ", b),
         x = "Observed x", y = "y") +
    theme_minimal(base_size = 11)
}

p1 <- make_plot(0,   "No noise (true x, \u03b2 = 1.5)")
p2 <- make_plot(0.5, "Low noise (sd = 0.5)")
p3 <- make_plot(1.5, "Medium noise (sd = 1.5)")
p4 <- make_plot(4,   "High noise (sd = 4)")

(p1 | p2) / (p3 | p4)
```

## Nunn & Qian: the potato and population {.smaller}

-   @nunn2011 ask: did the introduction of the potato to Europe after 1700 drive population growth?

-   The potato originated in South America; it spread through Europe unevenly, largely determined by **soil and climate suitability**

-   This suitability is plausibly unrelated to other determinants of population growth — which makes it a useful source of variation

-   Simplified regression:

$$y_{it} = \beta \cdot \text{PotatoSuitable}_i \times \mathbf{1}[\text{After 1700}]_t + \alpha_i + \tau_t + \epsilon_{it}$$

-   $y_{it}$: log population; $\alpha_i$: region fixed effects; $\tau_t$: period fixed effects

-   $\beta$ captures the extra population growth **after 1700** in regions more suitable for potatoes

## The data: McEvedy & Jones {.smaller}

-   Population data come from McEvedy & Jones (MJ), *Atlas of World Population History*

-   MJ describe their construction process openly.
    Key features per @guinnane2023

    -   Figures are **rounded** to the nearest unit (where the unit itself depends on the size of the population)
    -   Estimates for periods with little evidence rely on **perceived level of economic activity**
    -   Growth rates are set to **round numbers** (e.g. "doubling every century")
    -   Countries with thin evidence are **benchmarked against neighbours**

-   To use MJ for sub-national regions, researchers **allocate population by land area share** — a further step not in MJ

-   **Question for discussion**: given how this data was constructed, what would you expect the measurement error to look like?
    Is it likely to be classical?

## Bibliography
