---
title: "Measurement Error"
author: "Gabriel Mesevage"
format:
  revealjs:
    slide-number: c/t
    chalkboard:
      buttons: true
editor:
  markdown:
    wrap: sentence
bibliography: references.bib
---

## Today's plan {.smaller}

1.  Classical measurement error and why averaging works
2.  Systematic but uncorrelated bias: we can still recover $\beta$
3.  Non-classical (correlated) bias: destroys inference
4.  Measurement error in the explanatory variable: attenuation bias
5.  Application: population data, Nunn & Qian, and Guinnane's critique

## Classical measurement error {.smaller}

-   Imagine timing a book-toss with your phone: some people hit stop too early, some too late

-   Index each student's measurement $\tau_i$; the true duration is $\tau^*$:

$$\tau_i = \tau^* + \epsilon_i$$

-   If errors are **classical** — mean-zero, symmetric, uncorrelated with $\tau^*$:

$$\text{Avg}(\tau_i) = \frac{1}{N}\sum_i \tau_i = \tau^* + \text{Avg}(\epsilon_i) \xrightarrow{N \to \infty} \tau^*$$

-   **Key result**: averaging over many noisy measurements recovers the truth

-   This is why repeated measurement and averaging are so powerful in empirical science

## The experiment — round 1 {.smaller}

:::: {.columns}
::: {.column width="35%"}

I will toss a book. Time how long it stays in the air (seconds, to 2 decimal places) and enter your measurement in the spreadsheet.

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
#| message: false
#| warning: false
library(qrcode)
qr <- qr_code("https://docs.google.com/spreadsheets/d/1395nN5Ks0V34Qj6R0qJOrLEjd2exci4G7p43NNSgJC8/edit?usp=sharing")
plot(qr)
```

:::
::: {.column width="65%"}

```{=html}
<iframe src="https://docs.google.com/spreadsheets/d/1395nN5Ks0V34Qj6R0qJOrLEjd2exci4G7p43NNSgJC8/pubhtml?widget=true&amp;headers=false" width="100%" height="420px" frameborder="0"></iframe>
```

:::
::::

## Systematic bias — still useful {.smaller}

-   Suppose all students are systematically slow: average error is $b > 0$, not zero

-   But the bias is the **same** regardless of which hand I toss with

-   Let $D_i = 1$ for right-hand toss, $D_i = 0$ for left. Then:

$$\tau_i = D_i \tau^R + (1-D_i)\tau^L + b + \epsilon_i$$

Rearranging:

$$\tau_i = \underbrace{\tau^L + b}_{\alpha} + \underbrace{(\tau^R - \tau^L)}_{\beta} D_i + \epsilon_i = \alpha + \beta D_i + \epsilon_i$$

-   We still recover $\beta$ — the **difference** between groups — correctly

-   Only the intercept $\alpha$ absorbs the bias

-   **Intuition**: if bias is the same across groups, it cancels in the comparison

## The experiment — round 2 {.smaller}

:::: {.columns}
::: {.column width="35%"}

I will toss with my **right** then **left** hand. Record each timing and which hand. We will estimate the difference between hands.

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
#| message: false
#| warning: false
library(qrcode)
qr <- qr_code("https://docs.google.com/spreadsheets/d/1395nN5Ks0V34Qj6R0qJOrLEjd2exci4G7p43NNSgJC8/edit?usp=sharing")
plot(qr)
```

:::
::: {.column width="65%"}

```{=html}
<iframe src="https://docs.google.com/spreadsheets/d/1395nN5Ks0V34Qj6R0qJOrLEjd2exci4G7p43NNSgJC8/pubhtml?widget=true&amp;headers=false" width="100%" height="420px" frameborder="0"></iframe>
```

:::
::::

## Non-classical bias: inference breaks down {.smaller}

-   Now suppose the bias is **larger for higher tosses** — and I throw higher with my right hand

-   Let $b^R \neq b^L$; the bias is correlated with the treatment $D_i$:

$$\tau_i = \underbrace{\tau^L + b^L}_{\alpha} + \underbrace{(\tau^R - \tau^L) + (b^R - b^L)}_{\beta} D_i + \epsilon_i$$

-   $\hat\beta$ now conflates the **true difference** with the **difference in biases** — we cannot separate them

-   A large $\hat\beta = 1$ second is consistent with:
    -   A real 1-second gap, zero bias difference
    -   No real gap, a 1-second bias difference
    -   Any combination in between

-   **Lesson**: measurement error correlated with your explanatory variable destroys inference

## Measurement error in the explanatory variable {.smaller}

-   Now suppose the error is in $x$, not $y$: we observe $x_i = x^*_i + u_i$

-   **Classical error**: mean-zero, uncorrelated with $x^*_i$ and $\epsilon_i$

The OLS estimator gives:

$$\hat\beta = \beta \cdot \frac{\text{Var}(x^*_i)}{\text{Var}(x^*_i) + \text{Var}(u_i)}$$

-   The fraction is between 0 and 1: **attenuation bias** — the coefficient is pushed toward zero

-   The larger $\text{Var}(u_i)$ relative to $\text{Var}(x^*_i)$, the more the coefficient shrinks

-   A *constant* error (zero variance) causes no bias — only variance in the error matters

-   **Important**: this holds only for classical error; non-classical error in $x$ can bias in any direction

## Attenuation bias — adding noise to $x$ {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| message: false
#| warning: false
library(ggplot2)
library(patchwork)

set.seed(42)
n <- 200
x_true <- rnorm(n, 0, 1)
y <- 1.5 * x_true + rnorm(n, 0, 0.5)

make_plot <- function(noise_sd, label) {
  x_obs <- x_true + rnorm(n, 0, noise_sd)
  df <- data.frame(x = x_obs, y = y)
  b <- round(coef(lm(y ~ x, data = df))[2], 2)
  ggplot(df, aes(x, y)) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
    labs(title = label, subtitle = paste0("\u03b2\u0302 = ", b),
         x = "Observed x", y = "y") +
    theme_minimal(base_size = 11)
}

p1 <- make_plot(0,   "No noise (true x, \u03b2 = 1.5)")
p2 <- make_plot(0.5, "Low noise (sd = 0.5)")
p3 <- make_plot(1.5, "Medium noise (sd = 1.5)")
p4 <- make_plot(4,   "High noise (sd = 4)")

(p1 | p2) / (p3 | p4)
```

## Application: population data and the potato {.smaller}

-   @nunn2011 study the impact of the potato on European population after 1700

-   Simplified regression:

$$y_{it} = \beta \cdot \text{PotatoSuitable}_i \times \mathbf{1}[\text{After 1700}]_t + \alpha_i + \tau_t + \epsilon_{it}$$

-   Outcome $y_{it}$: log population, from McEvedy & Jones (MJ) *Atlas of World Population History*

-   MJ's construction: round to nearest $x$, guesses from perceived economic activity, round-number growth rates, neighbour benchmarks

-   To assign MJ figures to sub-national regions, researchers allocate by **land area share** — a further approximation

-   @guinnane2023 argues this measurement error is almost certainly **non-classical**

## Why non-classical? Guinnane's critique {.smaller}

-   MJ's errors are likely **larger in magnitude for larger populations** — more aggressive rounding for big countries

-   The errors are likely **correlated with potato suitability**: suitable areas are specific agricultural zones with particular population patterns

-   Area-based allocation of population introduces errors proportional to the size and economic character of the region — not random

Therefore:

-   The bias in $\hat\beta$ is **unknown in sign and magnitude**

-   Scholars who argue "the bias just attenuates" are invoking classical measurement error — an assumption that requires justification

-   Guinnane: the assumption is implausible given how the data were constructed

-   **Lesson**: always ask *how* your data were constructed before assuming measurement error is classical

## Bibliography
